---
title: Solving a differential equation numerically with R
author: 'Jordan '
date: '2019-04-06'
slug: solving-a-differential-equation-numerically-with-r
categories:
  - physics 
tags:
  - physics 
toc: no
images: ~
---

Today I am going to code a function that solves an ordinary differential equation numerically. The system I will be looking at is a pendulum. The pendulum with have a length equal to l, and a position of the pendulum can be described is l* theta. Theta is the angle of the string away from the vertical axis.

First I am going to add variables for physical constants 
```{r}
m <- 3 #Mass
g <- 9.82 #Gravitational constant 
l <- 2 #Length of the pendulum
mu <- .1 # Mu represents the loss of energy due to air resistance+
```

Next I'm going to create a function that takes arguments theta and theta_dot and returns theta_dubdot. m\*g/l is also known as the torque (thanks Tyler) and multiplying the torque by sin(theta) shows the relationship between the angle and the speed. Lastly the terms -mu\*theta_dot is showing the air resistance conditioned on the velocity. 
```{r}
theta_dubdot <- function(theta, theta_dot){
  return( -mu * theta_dot - (m*g/l) * sin(theta)) 
}

```

Now to set the starting conditions of theta and theta_dot

```{r}
theta0 <- pi/3 #60 degrees 
theta_dot0 <- 0 #Zero angular velocity 
```

Next I am going to code a solution to the equation that takes time as a input and returns theta. This function will approximate the theta at the end of the time t by taking a bunch of tiny steps equal to t/10.
```{r}
find_theta <- function(t.end) {
  theta <- theta0
  theta_dot <- theta_dot0
  delta.t <- .1
  for (t in seq(from = 0, to = t.end, by = delta.t)){
    theta <- theta + theta_dot * delta.t
    theta_dot <- theta_dot + theta_dubdot(theta, theta_dot) * delta.t
  }
  return(theta)
}

find_theta(6)
```

Now I want to create a for loop that stores the values for theta between t= 1 and t = 15

```{r}
results <- data.frame("t" = seq(0,15,.1), "theta" = seq(0,15,.1))
index <- 0
for (i in seq(0,15,.1)) {
  index <- index + 1
  results[index, ]$theta <- find_theta(i)
}

```

```{r include=FALSE}
library(ggplot2)
```


Lastly I am going to plot the pendulum position as theta*length over time. 

```{r fig.height=4.5, fig.width=6}
ggplot(results, aes(t,theta*l)) + 
  geom_line() + 
  theme_dark() + 
  xlab("time") + 
  ylab("Theta * Length")
 
```

***



***

## Now to visualize how changing starting positions can change the swing of the pendulum

After the interactive plot I will include the UI and Server code I used to make it. I used Shiny r. I hosted it for free on shinyapps.io and then embedded the app in to my blogdown rmd file.

<iframe width="850" height="700" scrolling="no" frameborder="no"  src="https://jordan-a.shinyapps.io/pendulum/"> </iframe>


```
# Load required packages --------------------------------------------------
library(shiny)
library(ggplot2)

# Build ui.R --------------------------------------------------------------
defaultTime <- 1  ## Changed it to 1 for rapid debugging
ui <- fluidPage(# Application title
  titlePanel("Physics engine pendulum."),
  # Sidebar with a slider input for mass
  sidebarLayout(
    sidebarPanel(
      sliderInput(
        "mass",
        "Mass:",
        min = 2,
        max = 50,
        value = 25
      ),
      sliderInput(
        "length",
        "Length:",
        min = 3,
        max = 50,
        value = 3
      ),
      sliderInput(
        "theta0",
        "Starting Theta:",
        min = 1,
        max = 180,
        value = 60
      ),
      sliderInput(
        "theta_dot0",
        "Starting Theta dot(Angular velocity):",
        min = 0,
        max = 5,
        value = 0
      ),
      sliderInput(
        "time",
        "How long do you want to observe the pendulum:",
        min = 1,
        max = 30,
        value = 15
      )
    ),
    
    # Show a plot of the generated distribution
    mainPanel(plotOutput("linePlot"))
  ))

# Build server.R ----------------------------------------------------------
server <- function(input, output) {
  #constants 
  g <- 9.82 #Gravitational constant
  mu <- .1 # Mu represents the loss of energy due to air resistance+
  
  # reactive function that finds theta dub dot
  theta_dubdot <- reactive(function(theta, theta_dot) {
    return(-mu * theta_dot - (input$mass * g / input$length) * sin(theta)) 
  })
  
  # reactive function that finds theta 
  find_theta <- reactive(function(t.end) {
    theta <- input$theta0
    theta_dot <- input$theta_dot0
    delta.t <- .1
    for (t in seq(from = 0, to = t.end, by = delta.t)) {
      theta <- theta + theta_dot * delta.t
      theta_dot <- theta_dot + theta_dubdot()(theta, theta_dot) * delta.t 
    }
    return(theta)
  })
  
  ### rv$result uses input$time so it needs to be reactive() not reactiveValues()
  rvResult <- reactive({
    req(input$time)
    outputDF <- data.frame("t" = seq(0, input$time , .1),
                           "theta" = seq(0, input$time , .1))
    index <- 0
    for (i in seq(0, input$time, .1)) {
      index <- index + 1
      computedTheta <- find_theta()(i)
      print(computedTheta)
      outputDF$theta[index] <- computedTheta 
    }
    return(outputDF)
  })
  
  output$linePlot <- renderPlot({
    print(rvResult())
    # draw the plot with the specified parameters
    ggplot(rvResult(), aes(t, theta * input$length)) +
      geom_line() +
      theme_dark() +
      xlab("Time") +
      ylab("Position") +
      labs(title = "Ploting the position of the weight over time")# creating the plot
  })
}

# Launch the Shiny app ----------------------------------------------------
shinyApp(ui = ui, server = server, options = list(launch.browser = TRUE))
```