---
title: 'A Coding Challenge: Alternating Case'
author: Jordan Arlan
date: '2019-06-24'
slug: a-coding-challenge-alternating-case
categories: []
tags: []
toc: no
images: ~
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
```

```{r message=FALSE, warning=FALSE}

# Complete the "toAlternatingCase" function below.  It should accept a
# character argument and return a character value.  The returned
# character value should be created by performing a transformation on
# the argument value in which:

# (1) Each odd-positioned alphabetical character is capitalized
# (2) Each even-positioned alphabetical character is made lowercase
# (3) Non-alphabetical characters are not changed in any way
#     [NOTE: the toupper() and tolower() methods which accept character
#      vectors do not modify non-alphabetical characters]

```

The goal is to create a robust function that passes all three rules set by the challenge. 

```{r}

toAlternatingCase <- function(str) {
  
  # First I am going to make all the letters lower case and vectorize the word. 
  # I make the letters lower case that way I know what I’m dealing with 
  # for the rest of the function.
  chr_vec <- tolower(str) %>%
    strsplit("") %>% unlist(use.names=FALSE)
  
  # Here I am saving the length of the string to use later
  l <- nchar(str)
  
  # In order to utilize the map_at function later on I will need a 
  # logical vector where TRUE corresponds to a desired upper case letter
  log_vec <- 1:l
  for (i in log_vec) {
    if ((i %% 2) == 0) {
      log_vec[i] <- FALSE
    } else {
      log_vec[i] <- TRUE
    }
  }
  # Now to actually manipulate the string to have an uppercase
  # letter at all odd positions. This will only be done if the 
  # length of the word is greater than one. 
  if(l > 1){
    chr_vec <- chr_vec %>% 
      map_if(as.logical(log_vec) == T, toupper)%>% unlist(use.names=FALSE)
  } else {
    chr_vec <- toupper(str)
  }
  #Lastly I combined the character vector back in to one word.
  chr_vec %>% paste0(collapse = "")
  
}
```
To test out the function I will run it a few times with some edge cases. 

```{r}
toAlternatingCase("Hello World")
toAlternatingCase("HELLO WORLD")
toAlternatingCase(" ")
toAlternatingCase("12345678")

```
## Unit Testing 

When it comes to building something like this, it is extremely helpful to always check your code as you go. I will show some examples of how I tested my code as figured the challenge out. 


```{r}
toAlternatingCase <- function(str) {
  chr_vec <- tolower(str) %>%
    strsplit("") %>% unlist(use.names=FALSE)
  print(c("this should be str seperated by letter in a vec", chr_vec))
  
  l <- nchar(str)
  print(c("this should be length of str", l))
  
  log_vec <- 1:l
  
  for (i in log_vec) {
    if ((i %% 2) == 0) {
      log_vec[i] <- FALSE
    } else {
      log_vec[i] <- TRUE
    }
  }
  print(c("this should be alternating T/F the length of str", log_vec))
}
```
Now to see what a test run looks like
```{r}
toAlternatingCase("Hello World")
```

